---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "FrankieZheng"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```


---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

**In this analysis, I look at median gross rent across counties in California, and
ask which communities we measure most and least reliably.**

---

## Choosing what to measure


```{r}
my_state <- "CA"        # two-letter abbreviation, in quotes
my_variable <- "B25064_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


**The dataset contains 58 rows, which matches the number of counties in California.**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```



Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

**The county with the largest margin of error is Sierra County, with a margin of error equal to 59.7% of its estimate.**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```



**There are 39 counties in the high-confidence category, 13 in moderate, and 6 in low confidence.**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```



**High-confidence counties have an average population of about 989117.821, while low-confidence counties have an average population of about 6770.667. This confirms that smaller counties tend to have less reliable data.**

---

## What this means for policy

In this analysis, I found that we know the least about small, rural counties in California. For example, Sierra County has a margin of error equal to 59.7% of its median gross rent estimate, meaning we cannot confidently say what the actual rent is. Trinity and Plumas counties also have high margins of error, above 14%. This pattern exists because the American Community Survey is a sample, not a full count. Smaller populations produce smaller sample sizes, which naturally result in greater statistical uncertainty. If a state agency allocated rental assistance funding using this data and ignored the margins of error, smaller counties like Sierra, Trinity, and Plumas could be severely hurt. Because their estimates are unreliable, they might be overlooked entirely, receiving far less funding than they actually need. This would exacerbate existing inequalities between urban and rural areas. Thus, ignoring uncertainty in ACS data can lead to inequitable policy outcomes, prioritizing well-measured, populous counties over the communities we know the least about.
